---
title: "Data Exploration"
author: "Mutsa Nyamuranga"
date: "2024-02-23"
output: html_document
---

# Setup

```{r}
#load needed packages. make sure they are installed.
library(here) #for data loading/saving
library(dplyr)
library(skimr)
library(ggplot2)
library(tidyr)
library(forcats)
library(gtsummary) #line added by Rachel for add_stat function to work
```

Load the data.

```{r}
#Path to data.
figdata <- here::here("data","processed_data","processeddata.rds")
#load data
explorfigdata <- readRDS(figdata)
```

# Data exploration through tables

Showing a bit of code to produce and save a summary table.



```{r}
# Calculate count and mean statistics
explorfigdata <- mutate_all(explorfigdata, as.numeric)
# Calculate count statistics
summary_stats <- explorfigdata %>%
  summarise(across(starts_with(c("Baseline", "Control", "BetaCaro")), list(Count = ~sum(!is.na(.))), .names = "{.col}_{.fn}"))

# Reshape the data for better presentation
summary_table <- summary_stats %>%
  pivot_longer(everything(), names_to = c(".value", "Variable"), names_sep = "_")
# Print the summary table
print(summary_table)
```

This table displays the counts of data per week. Where 16 would represent the number of weeks were data was collected for the Baseline variable mice. 

```{r}
# save to file
summary_table = here("results", "summarytable1.rds")
saveRDS(explorfigdata, file = summary_table)
```

```{r}
body_weights <- pivot_longer(explorfigdata, -Week, names_to = "Category", values_to = "BodyWeight")

summary_table2 <-
  tbl_summary(
    body_weights,
    by = Category,
    missing = "no" # don't list missing data separately
  ) %>%
  add_n() %>% # add column with total number of non-missing observations
  modify_header(label = "**Variable**") %>% # update the column header
  bold_labels()

summary_table2

# save to file
summarytable_elife = here("results", "summarytable.rds")
saveRDS(summary_table2, file = summarytable_elife)
```

With some final exploration. We look at the change in body weight over the 16 weeks. We see that males start and finish with higher body weights as expected. Of note, we see that the control and Beta-Carotene condition for female mice ends with the lowest body weights.

# Data exploration through figures

Here, I have created code that develops a bar graph to visualize the distribution of body weight in each condition. This will show the essential volume of body weight throughout the 16 weeks. This will not not be reliable for inferences but essential for drawing a baseline.

## Total Weights

```{r}
# Convert columns to numeric if needed
explorfigdata <- mutate_all(explorfigdata, as.numeric)

# Remove the "Week" variable
explorfigdata1 <- select(explorfigdata, -Week)

# Reshape the data and calculate sum for each condition
explorfigdata_long <- explorfigdata1 %>%
  gather(condition, value) %>%
  filter(!is.na(value)) %>%  # Remove NA values if any
  group_by(condition) %>%
  summarise(sum_value = sum(value))

# Create Bar Graph for Visualization
bar1 <- ggplot(explorfigdata_long, aes(x = fct_reorder(condition, desc(sum_value)), y = sum_value, fill = condition)) +
  geom_col() +
  scale_fill_manual(values = c("red", "blue", "blue", "red", "red", "blue")) +  # Fill colors
  scale_color_manual(values = c("red", "blue", "blue", "red", "red", "blue")) +  # Border colors
  labs(title = "Total Body Weight for Conditions",
       x = "Condition", y = "Sum of Weights")

plot(bar1)
# Saving the plot as an image file
figure_file = here("results","condition_distro1.png")
ggsave(filename = figure_file, plot=bar1) 
```

I color coded for male and female stratified conditions. Here, we see that the beta-carotene condition for males has the highest overall total body weight, while the baseline female condition has the lowest weight.

## Weight over time

```{r, message=FALSE, warning=FALSE}
# Reshape data to long format
library(tidyr)
body_weights_long <- pivot_longer(explorfigdata, -Week, names_to = "Category", values_to = "BodyWeight")

summary(body_weights_long)
str(body_weights_long)

# Plot the data
library(ggplot2)
line1 <- ggplot(body_weights_long, aes(x = Week, y = BodyWeight, color = Category)) +
  geom_line() +
  labs(title = "Comparison of Body Weight Changes Across Categories",
       x = "Week", y = "Body Weight") +
  theme_minimal()

plot(line1)

figure_file2 = here("results", "figures", "category_distro.png")
ggsave(filename = figure_file2, plot= line1) 
```
The line graph of total body weight over time reflects changes in body weight across different time points. Even if the sum of baseline body weights is lower, the trend over time may show an increase or decrease in body weight that is not evident when only considering the sum of baseline weights.
